Queue Aware Precoder Design for Space Frequency Resource Allocation

### 2014-08-13 Report/conclusions.tex
Finalized before reviewing - 2

In this work, we presented the \acl{QM} objective to minimize the number of backlogged packets by designing the transmit precoders across the space-frequency dimension in a multi-cell \ac{MU} \ac{MIMO} system. We proposed the \ac{JSFRA} scheme, which adpots the \ac{SCA} technique to model the nonconvex constraint as a convex constraint in an iterative manner to design the precoders for the \acl{QM} objective. We also proposed an alternative approach using the \ac{MSE} relaxation for the same problem when the receive beamformers are based on the \ac{MMSE} receivers. The exponent used in the proposed formulation can be modified to achieve either greedy allocation or equally fair allocation. We also proposed the distributed solutions for the proposed centralized \ac{JSFRA} problem using the \ac{ADMM} method, which has better convergence behaviour compared to the primal decomposition. In addition, we also proposed an iterative algorithm based on the \ac{KKT} conditions for the \ac{MSE} reformulated \ac{JSFRA} scheme to design the preocders in a distributed manner using the closed form solutions.

>>>>

In this work, we presented the \acl{QM} objective to minimize the number of backlogged packets by designing the transmit precoders across the space-frequency dimension in a multi-cell \ac{MU} \ac{MIMO} system. We proposed the \ac{JSFRA} scheme, which adopts the \ac{SCA} technique to model the nonconvex constraint as a convex constraint in an iterative manner to design the precoders for the \acl{QM} objective. We also proposed an alternative approach using the \ac{MSE} relaxation for the same problem when the receive beamformers are based on the \ac{MMSE} receivers. The exponent used in the proposed formulation can be modified to achieve either greedy allocation or equally fair allocation. We also proposed the distributed solutions for the proposed centralized \ac{JSFRA} problem using the \ac{ADMM} method, which has better convergence behaviour compared to the primal decomposition. In addition, we also proposed an iterative algorithm based on the \ac{KKT} conditions for the \ac{MSE} reformulated \ac{JSFRA} scheme to design the precoders in a distributed manner using the closed form solutions.

### 2014-08-13 Report/introduction.tex
Finalized before reviewing - 2

The last mile wireless connectivity poses significant bottleneck in the overall data traffic for the interconnected networks. The main challenges in the wireless networks are due to the scarcity of the available resources either in terms of power or spectrum usage and the complexity of the receiver algorithms which are proportional to the mobile battery drain. In order to overcome the receiver complexity, \ac{OFDM} based transmissions are introduced for the wideband transmissions. To improve data rate, multiple antennas are installed at \acp{BS} and/or at user terminals to avail additional freedom in the form of spatial dimension. The inclusion of \ac{MIMO} technique in wireless networks provides higher data rate or lower outage for the same transmission power and bandwidth.

@@

In this paper, we consider the problem of precoder design across the space-frequency resources to minimize the total number of queued packets of all \acp{BS}. For this highly nonconvex problem, we first propose two centralized methods. In the first method, we relax the nonconvex constraint by the first order taylor approximation around an operating point, which is updated in an iterative manner until convergence or to a certain accuracy. In the second method, we reformulate the \ac{JSFRA} problem using the \ac{MSE} equivalence with the rate expression to solve for the optimal precoders. For distributed implementation, we further proposed decentralized approaches based on primal and \ac{ADMM} scheme to identify the precoders independently across the \acp{BS} by exchanging limited information via backhaul. We also proposed an iterative algorithm by solving the \ac{KKT} equations, which can be implemented easily in a distributed manner.

>>>>

The last mile wireless connectivity poses significant bottleneck in the overall data traffic for the interconnected networks. The main challenges in the wireless networks are due to the scarcity of the available resources either in terms of power or spectrum usage and the complexity of the receiver algorithms which are proportional to the mobile battery drain. In order to overcome the receiver complexity, \ac{OFDM} based transmissions are introduced for the wide band transmissions. To improve data rate, multiple antennas are installed at \acp{BS} and/or at user terminals to avail additional freedom in the form of spatial dimension. The inclusion of \ac{MIMO} technique in wireless networks provides higher data rate or lower outage for the same transmission power and bandwidth.

@@

In this paper, we consider the problem of precoder design across the space-frequency resources to minimize the total number of queued packets of all \acp{BS}. For this highly nonconvex problem, we first propose two centralized methods. In the first method, we relax the nonconvex constraint by the first order taylor approximation around an operating point, which is updated in an iterative manner until convergence or to a certain accuracy. In the second method, we reformulate the \ac{JSFRA} problem using the \ac{MSE} equivalence with the rate expression to solve for the optimal precoders. For distributed implementation, we further proposed decentralized approaches based on primal and \ac{ADMM} scheme to identify the precoders independently across the \acp{BS} by exchanging limited information via back-haul. We also proposed an iterative algorithm by solving the \ac{KKT} equations, which can be implemented easily in a distributed manner.

### 2014-08-13 Report/kkt_mse_approach.tex
Finalized before reviewing - 2

In order to perform the distributed approach, we can consider the method proposed in \cite{komulainen2013effective}, where the users evaluate the \ac{MSE}, and the dual variable \me{\alpha_{l,k,n}} using the transmission made by all \acp{BS} using the transmit precoders \me{\mvec{m}{l,k,n}^{(i-1)}}. Once the dual variables are evaluated, the users will notify the dual variables and the receive beamformers to the \acp{BS} using the uplink precoded pilot signaling, where the precoders used for the uplink pilots are the receive beamformers \me{\mvec{\tilde{w}}{l,k,n}^{(i-1)} = \sqrt{\alpha_{l,k,n}^{(i-1)}}\mvec{w}{l,k,n}^{\ast(i-1)}} evaluated at the receivers. Upon receiving the uplink precoded pilots by the \ac{BS}, the effective channel \me{\mvec{H}{b,k,n}^{\mathrm{T}} \mvec{\tilde{w}}{l,k,n}^{(i-1)}} can be used in the expression \eqref{kkt-mse-4.3} to update the transmit precoders at the respective \acp{BS}, \me{\mbf{x}^{\ast}} represents the conjugate of \me{\mbf{x}}. The algorithmic representation of the distributed \ac{MSE}-\ac{KKT} scheme is shown in Algorithm. \ref{algo-4}. In this approach, the coordinating \acp{BS} are required to perform short transmissions to update the transmit and the receive beamformers before the actual data transmission.
\begin{algorithm}
 \SetAlgoLined
 \DontPrintSemicolon
 \BlankLine
 \SetKwInput{KwInit}{Initialize}
 \KwIn{\me{a_k, \, Q_k, \, \mvec{H}{b,k,n},\; \fall b \in \mathcal{B}, \, \fall k \in \mathcal{U}, \fall n \in \mathcal{C}}}
 \KwOut{\me{\mvec{m}{l,k,n}} and \me{\mvec{w}{l,k,n} \fall l \in \set{1,2,\dotsc,L}}}
 \KwInit{\me{i=1} and the receive beamformers \me{\mbf{w}^{(0)}_{l,k,n}} randomly}
 \KwInit{\me{\epsilon_{l,k,n}^{(0)}} randomly and the dual variables \me{{\alpha}_{l,k,n}^{(0)} = 1}}
 for each \ac{BS} \me{b \in \mc{B}}, perform the following procedure \;
 set the maximum iteration counter \me{I_{\max},J_{\max}} to a valid number. \;
 initialize \me{j = 0}. \;
 \Repeat{until convergence or \me{j \geq J_{\max}}}{
 initialize \me{i = 0}. \;
 \Repeat{until convergence or \me{i \geq I_{\max}}}{
	update the transmit precoders \me{\mvec{M}{k,n}^{(i)}} using \eqref{kkt-mse-4.3}, where \me{\delta_b} is identified by the bisection method satisfying \eqref{kkt-mse-1.4}. \;
    update \me{\epsilon_{l,k,n}^{(i)}}, \me{t_{l,k,n}^{(i)}}, \me{\sigma_{l,k,n}^{(i)}} and \me{\alpha_{l,k,n}^{(i)}} using \eqref{kkt-mse-4.4} and \eqref{kkt-mse-4.5}, \eqref{kkt-mse-4.2} and \eqref{kkt-mse-4.1} at the receivers using the transmit precoders \me{\mvec{M}{k,n}^{(i)}} from all \acp{BS}. \;
    perform precoded uplink pilot transmission from all users with the precoder \me{\mvec{\tilde{w}}{l,k,n}^{(j)}}. \;
    \me{i = i + 1} \;
  }
 evaluate the receive beamforming vector \me{\mvec{W}{k,n}^{(j+1)}} using \eqref{kkt-mse-4.6}. \;
 inform \acp{BS} using uplink precoded pilot transmission from all users with the precoder \me{\mvec{\tilde{w}}{l,k,n}^{(j+1)}}. \;
 \me{j = j + 1}. \;
 }
 \caption{\ac{KKT} approach for the \ac{JSFRA} scheme}
 \label{algo-4}
\end{algorithm}

>>>>

In order to perform the distributed approach, we can consider the method proposed in \cite{komulainen2013effective}, where the users evaluate the \ac{MSE}, and the dual variable \me{\alpha_{l,k,n}} using the transmission made by all \acp{BS} using the updated transmit precoders \me{\mvec{m}{l,k,n}^{(i-1)}}. Once the dual variables are evaluated, the users will notify the dual variables and the receive beamformers to the \acp{BS} using uplink precoded pilots, where the uplink precoder is given by \me{\mvec{\tilde{w}}{l,k,n}^{(i-1)} = \sqrt{\alpha_{l,k,n}^{(i-1)}}\mvec{w}{l,k,n}^{\ast(i-1)}}. Upon receiving the uplink precoded pilots at the \ac{BS} \me{b}, the effective channel \me{\mvec{H}{b,k,n}^{\mathrm{T}} \mvec{\tilde{w}}{l,k,n}^{(i-1)}} can be measured and used in the expression \eqref{kkt-mse-4.3} to update the transmit precoders, where \me{\mbf{x}^{\ast}} represents the conjugate of \me{\mbf{x}}. The algorithmic representation of the distributed \ac{MSE}-\ac{KKT} scheme is shown in the Algorithm. \ref{algo-4}. In this approach, the coordinating \acp{BS} are required to perform short transmissions to update the transmit and the receive beamformers until convergence or to a certain accuracy before doing the actual data transmission.
\begin{algorithm}
 \SetAlgoLined
 \DontPrintSemicolon
 \BlankLine
 \SetKwInput{KwInit}{Initialize}
 \KwIn{\me{a_k, \, Q_k, \, \mvec{H}{b,k,n},\; \fall b \in \mathcal{B}, \, \fall k \in \mathcal{U}, \fall n \in \mathcal{C}}}
 \KwOut{\me{\mvec{m}{l,k,n}} and \me{\mvec{w}{l,k,n} \fall l \in \set{1,2,\dotsc,L}}}
 \KwInit{\me{i=1} and the receive beamformers \me{\mbf{w}^{(0)}_{l,k,n}} randomly}
 \KwInit{\me{\epsilon_{l,k,n}^{(0)}} randomly and the dual variables \me{{\alpha}_{l,k,n}^{(0)} = 1}}
 for each \ac{BS} \me{b \in \mc{B}}, perform the following procedure \;
 set the maximum iteration counter \me{I_{\max},J_{\max}} to a valid number. \;
 initialize \me{j = 0}. \;
 \Repeat{until convergence or \me{j \geq J_{\max}}}{
 initialize \me{i = 0}. \;
 \Repeat{until convergence or \me{i \geq I_{\max}}}{
	update the transmit precoders \me{\mvec{M}{k,n}^{(i)}} using \eqref{kkt-mse-4.3}, where \me{\delta_b} is identified by the bisection method satisfying \eqref{kkt-mse-1.4}. \;
    update \me{\epsilon_{l,k,n}^{(i)}}, \me{t_{l,k,n}^{(i)}}, \me{\sigma_{l,k,n}^{(i)}} and \me{\alpha_{l,k,n}^{(i)}} using \eqref{kkt-mse-4.4} and \eqref{kkt-mse-4.5}, \eqref{kkt-mse-4.2} and \eqref{kkt-mse-4.1} at the receivers using the transmit precoders \me{\mvec{M}{k,n}^{(i)}} from all \acp{BS}. \;
    perform precoded uplink pilot transmission from all users with the precoder \me{\mvec{\tilde{w}}{l,k,n}^{(j)}}. \;
    \me{i = i + 1} \;
  }
 evaluate the receive beamforming vector \me{\mvec{W}{k,n}^{(j+1)}} using \eqref{kkt-mse-4.6}. \;
 inform \acp{BS} using uplink precoded pilot transmission from all users with the precoder \me{\mvec{\tilde{w}}{l,k,n}^{(j+1)}}. \;
 \me{j = j + 1}. \;
 }
 \caption{\ac{KKT} approach for the \ac{JSFRA} scheme}
 \label{algo-4}
\end{algorithm}

### 2014-08-13 Report/simulation_distributed.tex
update
Finalized before reviewing - 2

Fig. \ref{fig-d-3.1} compares the performances of the centralized algorithm based on the \ac{MSE} reformulation with the iterative approach proposed in Section \ref{sec-4.3} based on solving the \ac{KKT} conditions. In Fig. \ref{fig-d-3.1}, the plots are compared by the surplus number of backlogged packets at the end of each iteration or the \ac{SCA} update point. Fig. \ref{fig-d-3.1} shows that the \me{\ell_1} norm for \ac{JSFRA} scheme provides better performance over rest of the schemes due to its greedy objective. The \ac{KKT} approach for \me{\ell_1} norm is not defined due to the non-differentiability of the objective as discussed in the Section \ref{sec-4.3}, thereby performs the worst of all other approaches. The heuristic method used in the figure is obtained by forcing the dual variable \me{\sigma_{l,k,n}} in \eqref{kkt-mse-4.2} to \me{0} when the queue deviation is negative \me{Q_k - t_k < 0}, if not, then it will be the same as in \eqref{kkt-mse-4.2}. The additional condition can be justified due to the dropping of absolute value operator from the objective. It can be seen that the heuristic method is not converged to the optimal point but oscillates near the optimal point with the deviation determined by the factor \me{\rho} used in \eqref{kkt-mse-4.1}.
\begin{figure*}
\centering
\subfloat[][Queue deviation]{
\includegraphics[width=0.48\textwidth]{fig-9-2}
\label{fig-d-3.1}}
\hfill
\subfloat[][Sum rate performance]{
\includegraphics[width=0.48\textwidth]{fig-10}
\label{fig-d-3.2}}
\caption{Convergence plot for \me{\lbrace N,N_B,K,N_R \rbrace = \lbrace 5,2,8,1 \rbrace} model}
\label{fig-d-3}
\end{figure*}

The objective values are mentioned in the legend for all the schemes, since the objective of \me{\ell_2} norm is not the same as \me{\ell_1} norm, which is used for the plot. The \me{\ell_2} norm for the \ac{JSFRA} and the \ac{KKT} based approach achieves nearly the same objective value of \me{6.62} but different \me{\chi}, since the dual variables are not allowed to converge in between the \ac{SCA} steps in the \ac{KKT} approach. In the simulation, we update all the variables at each iteration due to the practical issues in the signaling between the \acp{BS}. The figure compares the effect of dropping the squared rate variable from the objective in the \ac{Q-WSRM} scheme compared to the \me{\ell_2} norm which includes it. By dropping the squared rate variable, the \ac{Q-WSRM} scheme minimizes the number of queued packets in a prioritized manner based on the respective queues. Contrastingly, the \me{\ell_2} norm allocate rates to the users with the higher number of queued packets before addressing the users with the smaller number of queued packets.

>>>>

Fig. \ref{fig-d-3.1} compares the performances of the centralized algorithm based on the \ac{MSE} reformulation with the iterative approach proposed in Section \ref{sec-4.3} based on solving the \ac{KKT} conditions. In Fig. \ref{fig-d-3.1}, the plots are compared by the surplus number of backlogged packets at the end of each iteration or the \ac{SCA} update point. Fig. \ref{fig-d-3.1} shows that the \me{\ell_1} norm for \ac{JSFRA} scheme provides better performance over rest of the schemes due to its greedy objective. The \ac{KKT} approach for \me{\ell_1} norm is not defined due to the non-differentiability of the objective as discussed in the Section \ref{sec-4.3}, thereby performs the worst of all other approaches. The heuristic method used in the figure is obtained by forcing the dual variable \me{\sigma_{l,k,n}} in \eqref{kkt-mse-4.2} to \me{0} when the queue deviation is negative \me{Q_k - t_k < 0}, if not, then it will be the same as in \eqref{kkt-mse-4.2}. The additional condition can be justified due to the dropping of absolute value operator from the objective. It can be seen that the heuristic method is not converged to the optimal point but oscillates near the optimal point with the deviation determined by the factor \me{\rho} used in \eqref{kkt-mse-4.1}.
\begin{figure}
\centering
\includegraphics[width=\columnwidth]{fig-9-2}
\caption{Convergence plot for \me{\lbrace N,N_B,K,N_R \rbrace = \lbrace 5,2,8,1 \rbrace} model}
\label{fig-d-3.1}
\end{figure}

The objective values are mentioned in the legend for all the schemes, since the objective of \me{\ell_2} norm is not the same as \me{\ell_1} norm, which is used for the plot. The \me{\ell_2} norm for the \ac{JSFRA} and the \ac{KKT} based approach achieves nearly the same objective value of \me{6.62} but different \me{\chi}, since the dual variables are not allowed to converge in between the \ac{SCA} steps in the \ac{KKT} approach. In the simulation, we update all the variables at each iteration due to the practical issues in the signaling between the \acp{BS}. The figure compares the effect of dropping the squared rate variable from the objective in the \ac{Q-WSRM} scheme compared to the \me{\ell_2} norm which includes it. By dropping the squared rate variable, the \ac{Q-WSRM} scheme minimizes the number of queued packets in a prioritized manner based on the respective queues. On contrary, the \me{\ell_2} norm allocate rates to the users with the higher number of queued packets before addressing the users with the smaller number of queued packets.

### 2014-08-13 Report/wsrm_formulation.tex
Finalized before reviewing - 2

Now, the optimal linear receiver for the fixed transmit precoders \me{\mbf{M}_{i,n} \, \forall i \in \mc{U}, \, \forall n \in \mc{C}} is obtained by minimizing \eqref{eqn-3} w.r.t \me{\mbf{w}_{l,k,n}}, which is given by
\begin{IEEEeqnarray}{rCl}\label{eqn-9--1}
\underset{\substack{\gamma_{l,k,n},p_{l,k,n},q_{l,k,n} \\ \mvec{w}{l,k,n}, \beta_{l,k,n}}}{\text{minimize}} &\quad & \| \tilde{\mbf{v}} \|_q \IEEEyessubnumber\label{eqn-9--1.1a} \\
\text{subject to} & \quad & \eqref{eqn-9.1c}, \eqref{eqn-9.1d}, \eqref{eqn-9.1e}, \text{ and } \eqref{eqn-8} \IEEEyessubnumber \label{eqn-9--1.1b}
\end{IEEEeqnarray}
Now, by solving for the optimal receiver using \ac{KKT} conditions, we obtian the following expression for the receive beamformer \me{\mvec{w}{l,k,n}} in an iterative manner as
\begin{subeqnarray} \label{opt-rx}
\mvec{R}{l,k,n} &=& \displaystyle \sum_{(j,i)\neq (l,k)} \mvec{H}{b_i,k,n} {\mbf{m}}_{j,i,n} {\mbf{m}}_{j,i,n}^\herm \mvec{H}{b_i,k,n}^\herm + N_0 \, \mathbf{I}_{N_R} \\
\mvec{w}{l,k,n}^{(i)} &=& \left ( \frac{\tilde{\beta}_{l,k,n} {\mbf{m}}_{l,k,n}^\herm \mvec{H}{b_k,k,n}^\herm \mbf{w}_{l,k,n}^{(i-1)} }{\|\mvec{w}{l,k,n}^{(i-1)} \mvec{H}{b_k,k,n} {\mbf{m}}_{l,k,n} \|^2} \right ) \, \mvec{R}{l,k,n}^{-1} \, \mvec{H}{b_k,k,n} {\mbf{m}}_{l,k,n}
\end{subeqnarray}
where \me{\mvec{w}{l,k,n}^{(i-1)}} denotes the receive beamformer found in the earlier iteration, around which the linear relaxation is performed to convexify the problem in \eqref{eqn-9--1}. It can be seen that the optimal receiver expression in \eqref{opt-rx} is the scaled version of the \ac{MMSE} receiver, which is given by
\begin{IEEEeqnarray}{rCl}\label{eqn-10}
\mvec{w}{l,k,n} &=& \mathbf{R}^{-1}_{l,k,n} \; \mvec{H}{b_k,k,n} \; \mvec{m}{l,k,n} \IEEEyessubnumber \\
\mvec{R}{l,k,n} &=& \displaystyle \sum_{i\in \mc{U}} \sum_{j=1}^L \mvec{H}{b_i,k,n} \mvec{m}{j,i,n} \mvec{m}{j,i,n}^\herm \mvec{H}{b_i,k,n}^\herm + N_0 \, \mathbf{I}_{N_R} \IEEEyessubnumber
\end{IEEEeqnarray}

>>>>

Now, the optimal linear receivers for the fixed transmit precoders \me{\mbf{M}_{i,n} \, \forall i \in \mc{U}, \, \forall n \in \mc{C}} are obtained by minimizing \eqref{eqn-3} w.r.t \me{\mbf{w}_{l,k,n}} as
\begin{IEEEeqnarray}{rCl}\label{eqn-9--1}
\underset{\substack{\gamma_{l,k,n},p_{l,k,n},q_{l,k,n} \\ \mvec{w}{l,k,n}, \beta_{l,k,n}}}{\text{minimize}} &\quad & \| \tilde{\mbf{v}} \|_q \IEEEyessubnumber\label{eqn-9--1.1a} \\
\text{subject to} & \quad & \eqref{eqn-9.1c}, \eqref{eqn-9.1d}, \eqref{eqn-9.1e}, \text{ and } \eqref{eqn-8} \IEEEyessubnumber \label{eqn-9--1.1b}
\end{IEEEeqnarray}
Solving \eqref{eqn-9--1} using \ac{KKT} conditions, we obtain the following iterative expression for the receive beamformer \me{\mvec{w}{l,k,n}} as
\begin{subeqnarray} \label{opt-rx}
\mvec{R}{l,k,n} &=& \displaystyle \sum_{(j,i)\neq (l,k)} \mvec{H}{b_i,k,n} {\mbf{m}}_{j,i,n} {\mbf{m}}_{j,i,n}^\herm \mvec{H}{b_i,k,n}^\herm + N_0 \, \mathbf{I}_{N_R} \\
\mvec{w}{l,k,n}^{(i)} &=& \left ( \frac{\tilde{\beta}_{l,k,n} {\mbf{m}}_{l,k,n}^\herm \mvec{H}{b_k,k,n}^\herm \mbf{w}_{l,k,n}^{(i-1)} }{\|\mvec{w}{l,k,n}^{(i-1)} \mvec{H}{b_k,k,n} {\mbf{m}}_{l,k,n} \|^2} \right ) \, \mvec{R}{l,k,n}^{-1} \, \mvec{H}{b_k,k,n} {\mbf{m}}_{l,k,n}
\end{subeqnarray}
where \me{\mvec{w}{l,k,n}^{(i-1)}} is the receive beamformer from the earlier iteration, upon which the linear relaxation is performed for the nonconvex constraint in \eqref{eqn-9--1}. It can be seen that the optimal receiver expression in \eqref{opt-rx} is the scaled version of the \ac{MMSE} receiver, which is given by
\begin{IEEEeqnarray}{rCl}\label{eqn-10}
\mvec{w}{l,k,n} &=& \mathbf{R}^{-1}_{l,k,n} \; \mvec{H}{b_k,k,n} \; \mvec{m}{l,k,n} \IEEEyessubnumber \\
\mvec{R}{l,k,n} &=& \displaystyle \sum_{i\in \mc{U}} \sum_{j=1}^L \mvec{H}{b_i,k,n} \mvec{m}{j,i,n} \mvec{m}{j,i,n}^\herm \mvec{H}{b_i,k,n}^\herm + N_0 \, \mathbf{I}_{N_R} \IEEEyessubnumber
\end{IEEEeqnarray}
